---
title: "Graphing migration in Puerto Rico 2015-2019 with IPUMS USA PRCS data"
author: "Your Name Here"
date: '2022-04-05'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
suppressPackageStartupMessages({
  library(ipumsr)
  library(tidyverse)
})
```

```{r load-data, eval=FALSE}
if (!file.exists("prcs_migration_extract.xml")) {
  # Load extract definition from JSON
  prcs_migration_extract <- define_extract_from_json(
    "prcs_migration_extract.json",
    "usa"
  )
  # Submit, wait for, and download extract
  ddi_filename <- submit_extract(prcs_migration_extract) %>% 
    wait_for_extract() %>% 
    download_extract() %>% 
    basename()
  # Infer data file name from DDI file name
  data_filename <- str_replace(ddi_filename, "\\.xml$", ".dat.gz")
  # Standardize DDI and data file names #<<
  file.rename(ddi_filename, "prcs_migration_extract.xml") #<<
  file.rename(data_filename, "prcs_migration_extract.dat.gz") #<<
}
```

```{r prep-data}
# Prep education variable
college_regex <- "^[123] year(s)? of college$"
data <- data %>% 
  mutate(
    EDUCD3 = EDUCD %>%
      lbl_collapse(~.val %/% 10) %>% 
      lbl_relabel(
        lbl(2, "Less than High School") ~.val > 0 & .val < 6,
        lbl(3, "High school") ~.lbl == "Grade 12", #<<
        lbl(4, "Some college") ~str_detect(.lbl, college_regex), #<<
        lbl(5, "College or more") ~.val %in% c(10, 11)
      ) %>%
      as_factor()
  )

# Prep income variable
value_to_quintile <- function(x) {
  cut_points <- quantile(x, probs = c(0.2, 0.4, 0.6, 0.8), na.rm = TRUE)
  cut(
    x, 
    breaks = c(-Inf, cut_points, Inf), 
    labels = c("Lowest", "Lower Middle", "Middle", "Upper Middle", "Highest"),
    ordered_result = TRUE
  )
}

hhincome_quintiles <- data %>% 
  filter(PERNUM == 1 & HHINCOME != 9999999) %>% 
  group_by(YEAR) %>% 
  mutate(hhincome_quintile = value_to_quintile(HHINCOME)) %>% 
  ungroup() %>% 
  select(YEAR, SERIAL, hhincome_quintile)

data <- data %>% 
  left_join(hhincome_quintiles, by = c("YEAR", "SERIAL"))

# Prep migration variable
data <- data %>% 
  mutate(
    moved_in_last_year = case_when(
      MIGRATE1 %in% c(0, 9) ~ NA, 
      MIGRATE1 == 1 ~ FALSE, 
      MIGRATE1 %in% 2:4 ~ TRUE
    )
  )


# Prep age variable
age_to_age_group <- function(x) {
  cut_points <- c(9, 17, 34, 64)
  cut(
    x,
    breaks = c(-Inf, cut_points, Inf),
    labels = c("0-9", "10-17", "18-34", "35-64", "65+"),
    ordered_result = TRUE
  )
}

data <- data %>% 
  mutate(age_group = age_to_age_group(AGE))
```

# Migration 2015-2019 {.tabset}

The percentage of people who had moved in the last year increased between 2017 
and 2018 from about 6% to over 8% among all persons in Puerto Rico, but the 
magnitude of this trend varies by education, household income, and age.

Note: These graphs show trends in point estimates from sample data, without 
displaying estimates of sampling error. Differences over time or across groups 
may not be statistically significant. To calculate confidence intervals for 
point estimates, follow the 
[IPUMS USA instructions for using replicate weights](https://usa.ipums.org/usa/repwt.shtml).

## Overall

```{r migration-graph-1, dpi=300, fig.height=5, fig.width=8, echo = FALSE}
data %>% 
  group_by(YEAR) %>% 
  summarize(
    pct_moved = 100 * sum(PERWT[moved_in_last_year]) / sum(PERWT)
  ) %>% 
  ggplot(aes(x = YEAR, y = pct_moved)) +
    geom_line() +
    labs(
      title = "Percentage of people who moved in the past year, 2015-2019",
      x = NULL,
      y = "%"
    )
```

## By education

```{r migration-graph-2, dpi=300, fig.height=5, fig.width=8, echo = FALSE}
data %>% 
  filter(AGE >= 25) %>%
  group_by(YEAR, EDUCD3) %>% 
  summarize(
    pct_moved = 100 * sum(PERWT[moved_in_last_year]) / sum(PERWT),
    .groups = "drop"
  ) %>% 
  ggplot(aes(x = YEAR, y = pct_moved)) +
    geom_line() +
    facet_wrap(~EDUCD3) +
    labs(
      title = "Percentage of people who moved in the past year, 2015-2019",
      subtitle = "Among persons age 25 and older",
      x = NULL,
      y = "%"
    )
```

## By household income

```{r migration-graph-3, dpi=300, fig.height=5, fig.width=8, echo = FALSE}
data %>% 
  filter(!is.na(hhincome_quintile)) %>% 
  group_by(YEAR, hhincome_quintile) %>% 
  summarize(
    pct_moved = 100 * sum(PERWT[moved_in_last_year]) / sum(PERWT),
    .groups = "drop"
  ) %>% 
  ggplot(aes(x = YEAR, y = pct_moved)) +
    geom_line() +
    facet_wrap(~hhincome_quintile) +
    labs(
      title = "Percentage of people who moved in the past year, 2015-2019",
      x = NULL,
      y = "%"
    )
```


## By age

```{r migration-graph-4, dpi=300, fig.height=5, fig.width=8, echo = FALSE}
data %>% 
  group_by(YEAR, age_group) %>% 
  summarize(
    pct_moved = 100 * sum(PERWT[moved_in_last_year]) / sum(PERWT),
    .groups = "drop"
  ) %>% 
  ggplot(aes(x = YEAR, y = pct_moved)) +
    geom_line() +
    facet_wrap(~age_group) +
    labs(
      title = "Percentage of people who moved in the past year, 2015-2019",
      x = NULL,
      y = "%"
    )
```